Synthesizing Quotient Lenses

### 2018-03-09 papers/QuotientLenses/main.tex
A few comments on comments

% FOR Regular Expression names
\newcommand{\re}[1]{\ensuremath{\mathtt{#1}}}
\newcommand{\kw}[1]{\ensuremath{\mathit{#1}}}
\newcommand{\codefont}[1]{\ensuremath{\mathsf{#1}}}
\newcommand{\project}[2]{\ensuremath{\kw{project} \; #1 \mapsto #2}}
\newcommand{\squash}[3]{\ensuremath{\kw{squash} \; #1 \rightarrow #2\; \kw{using} \; #3}}
\newcommand{\perm}[2]{\ensuremath{\kw{perm}\; (#1)\; \kw{with}\; #2}}
\newcommand{\normalize}[3]{\ensuremath{\kw{normalize} \; (#1, #2, #3)}}
\newcommand{\sep}{\ensuremath{\ | \ }}
\newcommand{\canonizer}{\ensuremath{\kw{canonizer}}}
\newcommand{\bibtex}{\textsc{Bib}\TeX{}}
\newcommand{\get}{\ensuremath{\kw{get}}}
\newcommand{\semicolon}{\ensuremath{\; ; \;}}
\newcommand{\lput}{\ensuremath{\kw{put}}}
\newcommand{\create}{\ensuremath{\kw{create}}}
\newcommand{\eqrel}[1]{\ensuremath{\equiv_{#1}}}

@@

\sam{Here is my response to this. You observe that the idea of using half a lens
as a canonizer comes out of nowhere, and that this ideas is a hack and always
was. I don't think that's quite right. I think that it is useful to have a
combinator which chooses between two or more formats. Indeed, the permutation
combinator is just a special case of this. The bibtex name field is a good
example of where such a combinator would be useful.  But if we choose to have
such a combinator, then we need a way of mapping one format to another, and
really the only interesting string-to-string transformations in Boomerang are
given by lenses, which is why we use the get function of a lens in a canonizer.
We could have used any other function between the two formats but Boomerang
doesn't give us too many options for writing unidirectional transformations. So
I don't think that using a lens as a canonizer is a trick. I think that
Boomerang just isn't designed for unidirectional tranformations. But I also
agree that it doesn't make much sense to synthesize a lens only to get rid of
one half of it.
I'm not sure what the solution is to be honest.}
\bcp{Maybe just admit a bit more frankly that this is a convenient trick
  to avoid introducing a whole new language for writing (or synthesizing)
  unidirectional functions.}

@@

(* allow any permutation of fields interspersed with arbitrary whitespace *)
let bib_fields = perm (label, bib_author, bib_title) with (project (wsp+ . "," . wsp+) $\to$ ",\n")
let bibtex = "@book{" . bib_fields . "}"
\end{lstlisting}
\caption{QRE specification of \bibtex{} records. }
\label{fig:example-qre}
\end{figure}
\subsection{QRE Lenses and QRE Lens Synthesis}
\label{sec:examplesynth}
At this point we already have a tool for synthesizing bijective lenses from a
pair of regular expressions and a set of example input-output pairs, as well as
the language of QREs for defining equivalence relations on regular expressions.
An obvious next step in simplifying the quotient lens programming is thus to use
the bijective lens synthesis procedure as a subroutine for a quotient lens
synthesis procedure. This new procedure should take two QREs and a set
of example input-output pairs as input, compute the canonical data formats on
both sides of the lens from the pair of QREs, map the example input-output pairs
to their canonical representatives to get a set of examples in canonical form,
and then invoke the bijective lens synthesis procedure on the canonical data
formats and the new set of examples that are in canonical form.

Indeed this basic idea is what motivates our definition of {\em QRE Lenses}.
Intuitively, just like Boomerang's quotient lenses, our QRE lenses are bijective
lenses with ``canonizers at the edges''---Figure~\ref{fig:attheedges} presents
the architecture of each lens.
\begin{figure}[t]
\centering
\includegraphics[width=0.7\textwidth]{canonizers-outside}
\caption{Quotient Lens with ``Canonizers at the Edges''
\sam{We don't need a general choose function. With QREs the kernel language is
always contained in the whole language so we just use the inclusion as the
choose function.}\bcp{Done.  But I wonder if the picture should mention QREs
explicitly, somehow... any ideas?}}\sam{Maybe we can put all structures, the
canonizer and the canonical strcutre in a box that says QRE?}\bcp{Like this?
 (But maybe we should go a step further and label things like $K(c)$,
 $W(c)$, Canonizer(c), etc.  And probably we should display K as a subset of
 W.  I can make these changes later.)}
\label{fig:attheedges}
\end{figure}
Every QRE lens $q$ has a type $q: c_1 \Leftrightarrow c_2$ where $c_1$ and $c_2$
are QREs. In the get direction, a QRE lens $q: c_1 \Leftrightarrow c_2$ uses the
source QRE $c_1$ to compute a unique representative for the data modulo the
equivalence relation defined by $c_1$ and then applies the $\get$ function of a
bijective lens $\ell$ to this representative. In the backward direction, $q$
operates similarly, but using the view QRE $c_2$ and the $\lput$ function of
$\ell$.

Because the QREs $c_1$ and $c_2$ determine the internal formats for data after
canonization, and because the algorithm for synthesizing bijective lenses is
directed by these formats, $c_1$ and $c_1$ are all that is required to
synthesize QRE lenses end-to-end. Indeed, we structured QRE lenses as bijective
lenses with canonizers at the edges, precisely to facilitate the use of this
synthesis algorithm.

@@

At this point we also bring to attention the fact that in order for the {\em
normalize} QRE to be well-formed, then canonizing function $f$ used by this
combinator must be surjective and idempotent. Unfortunately, verifying
that a function satisfies these properties is in general undecidable.
Nonetheless, we allow a programmer to use normalize at their own risk without
checking these side conditions.

We do this firstly because the {\em normalize} combinator allows one to express
equivalence relations that are either difficult to define, or that cannot be
defined using the other combinators. Secondly, the {\em normalize} canonizer
enjoys the property that each of the other combinators $c$ is semantically
equivalent to $\normalize{W(c)}{K(c)}{f}$ for some surjective, idempotent
function $f : \mathcal{L}(W(c)) \longrightarrow \mathcal{L}(K(c))$. The {\em
normalize} combinator hence gives a sufficient condition for potential QREs to
be ``well-behaved'', particularly with respect to the strategy for synthesizing
quotient lenses which we shall introduce in Section~\ref{synth}.  \bcp{I
  think here we're mixing concerns of language design with the concerns of
  the underlying theory.  Indeed, I think it would be clearer to reorganize
  the whole section as follows: first present just normalize and discuss its
  properties; then observe that the side condition is annoying because it
  cannot be checked automatically; then introduce project, squash, and perm
  as a pragmatically motivated, extensible set of safe combinators that
  we've found useful for a broad range of examples.  This will also help
  address the complaint that extracting half of a lens is a hack.}

In spite of this ``completeness'' property of the {\em normalize} combinator,
our experience with QREs shows that using the {\em project}, {\em squash} and
{\em perm} combinators as well as the regular operators applied to these is
much easier than using just {\em normalize} because of the compositional nature
of the other combinators.  \bcp{Maybe this comment is not needed?  It seems
  obvious, besides the fact that these other combinators are safe while
  normalize requires a proof by hand.}

The formal semantics of QREs satisfy the following theorem:\bcp{That bit is
  entirely redundant with the theorem itself.  Either simply delete it or
  replace with something more interesting. :-)}
\begin{theorem}
If $c$ is a well formed QRE, then
\begin{enumerate}
  \item $W(c)$ and $K(c)$ are well defined regular expressions,
  \item  $\eqrel{c}$ is an equivalence relation on $\mathcal{L}(W(c))$,
  \item  $\mathcal{L}(K(c))$ forms a complete set of representatives for
  $\eqrel{c}$,
  \item $\canonizer(c):\mathcal{L}(W(c)) \longrightarrow \mathcal{L}(K(c))$ is
  a well-defined function, and
  \item  given any $w \in \mathcal{L}(W(c))$, $\canonizer(c)(w)$ is the unique
  $k$ in $\mathcal{L}(K(c))$ such that $k$ is equivalent to $w$ modulo
  $\eqrel{c}$.
\end{enumerate}
\end{theorem}

@@

\bcp{A question about variable names: If we use $R$ and $S$ for plain
  regular expressions, it seems inconsistent to use $c$ for QREs.  (And even
  more inconsistent to use $c$ and $c'$ or $c$ and $c_1$!) I suggest $P$ and
  $Q$ for QREs (with $Q$ as the default when we are only talking about
  one).}

Given regular expressions $R, S$ and equivalence relations
$\sim_R$ and $\sim_S$ defined on $\mathcal{L}(R)$ and $\mathcal{L}(S)$
respectively, we\bcp{can we clarify that this is exactly Foster et al's definition?} define a {\em bijective quotient lens} $q : R /{\sim_R}
\Leftrightarrow S/{\sim_S}$ to be a pair of functions $q.get :
\mathcal{L}(R) \longrightarrow \mathcal{L}(S)$ and $q.put : \mathcal{L}(S)
\longrightarrow \mathcal{L}(R)$ such that for all $r \in \mathcal{L}(R)$ and $s
\in \mathcal{L}(S)$,
\begin{align*}
q.put(q.get(r)) &\sim_R r \text{, and}\\
q.get(q.put(s)) &\sim_R s
\end{align*}
Moreover, if $r \sim_R r'$, then $q.get(r) = q.get(r')$, and if $s \sim_S s'$,
then $q.put(s) = q.put(s')$.  \bcp{We should emphasize that this is
  precisely the desired loosening of the too-strict plain lens laws.}

In words, a bijective quotient lens $q$ from $R$ to $S$ modulo $\sim_R$
and $\sim_S$ is a pair of functions $q.get$ and $q.put$ such that $q.get$
respects $\sim_R$ and $q.put$ respects $\sim_S$, and such that the $get$ and
$put$ functions lifted to $\mathcal{L}(R)/{\sim_R}$ and
$\mathcal{L}(S)/{\sim_S}$ are mutual inverses.

\bcp{Typesetting: $q.get$ and $q.\get$ are subtly different, as are
  $q.put$ and $q.\lput$, and the latter forms are the correct ones.
  In general, you should never put a whole word in math mode without
  wrapping it in mathit: the inter-letter spacing will be wrong.}

Now let $c$ and $c_1$ be QREs. Given a bijection $\ell : K(c) \Leftrightarrow
K(c_1)$, we may define a quotient lens $q : W(c)/\eqrel{c} \Leftrightarrow
W(c_1)/\eqrel{c_1}$ by
\begin{equation}\label{normalform}
q.get = \ell \circ \canonizer(c) \text{ and } q.put = \ell^{-1} \circ
\canonizer(c_1),
\end{equation}

\noindent In other words, the get function of the $q$ first canonizes the source
data using $\canonizer(c)$ and then applies $f$ to the result, while the
put function of the $q$ first canonizes the target data using
$\canonizer(c_1)$ and the applies $f^{-1}$ to the result. This is our approach in
defining QRE lenses.

One potential pitfall with QRE lenses is that the combinators that
we define on them may not preserve this normal form. For example, the
composition $q' \; ; \; q$ of QRE lenses $q$ and $q'$ may not be equivalent to
any quotient lens that has the canonical normal form of a bijective lens with
canonizers a the edges. Consequently, in order to guarantee that every QRE
lens $q$ is equivalent to a QRE lens $q'$ which has the canonical form of
a bijection with canonizers at the edges, we need to identify a set $S$ of
bijections that will ensure that QRE lenses are closed under the action of QRE
lens combinators. Our choice is to use the class of \textit{bijective lenses},
which we briefly discuss in the next section.

\subsection{Bijective Lenses}
We define the set of \textit{bijective lenses} to be the set of bijections
between regular languages created using Boomerang lens combinators.
The syntax for the language of bijective lenses is given by
$$\ell := \mathit{id} \; (R) \sep const(s, s') \sep  swap(\ell,
\ell) \sep \ell \cdot \ell' \; |  \; (\ell \sep \ell') \sep \ell^* \;
| \; \ell \circ \ell',$$ where $R$ ranges over regular expressions and $s$
ranges over character strings.

The denotation of a lens $\ell$ is $\llbracket \ell \rrbracket \subseteq
\mathit{String} \times \mathit{String}$. If $(s_1, s_2) \in \llbracket \ell
\rrbracket$, then $\ell$ maps between $s_1$ and $s_2$.
\begin{align*}
\llbracket R \rrbracket &= \{(r, r) \sep r \in \mathcal{L}(R)\}\\
\llbracket const(r, s) \rrbracket &= \{(r, s)\}\\
\llbracket swap(\ell, \ell') \rrbracket &= \{(s \cdot t, t' \cdot s') \sep
(s, s') \in \llbracket \ell \rrbracket \text{ and } (t, t') \in \llbracket
\ell' \rrbracket\}\\
\llbracket \ell \cdot \ell' \rrbracket &= \{(s \cdot t, s' \cdot t) \sep
(s, s') \in \llbracket \ell \rrbracket \text{ and } (t, t') \in \llbracket
\ell' \rrbracket\}\\
\llbracket \ell \sep \ell' \rrbracket &= \{(s \cdot t) \sep
(s, t) \in \llbracket \ell \rrbracket \text{ or } (s, t) \in \llbracket
\ell' \rrbracket\}\\
\llbracket \ell^* \rrbracket &= \{(s_1 \cdot \ldots \cdot s_n, t_1 \cdot \ldots
\cdot t_n) \sep (s_i, t_i) \in \llbracket \ell \rrbracket \text{ for } 1
\leq i \leq n\}\\
\llbracket \ell_2 \circ \ell_1 \rrbracket &= \{(r, t) \sep \text{there exists }s
\text{ such that } (r, s) \in \llbracket \ell_1 \rrbracket \text{ and } (s, t)
\in \llbracket \ell_2 \rrbracket\}
\end{align*}
The $\mathit{const}(s, t)$ lens replaces the string $s$ with $t$ in the source
data in the forward direction, and $t$ with $s$ in the target data in the backward
direction. $\mathit{id}(R)$ applies the identity function to the source and target
in $\mathcal{L}(R)$ in both directions. The composition lens $\ell' \circ \ell$
applies $\ell$ followed by then $\ell'$ to the source in the forward direction,
and applies $\ell'$ followed by $\ell$ to the target in the backward direction.
The lens $\ell \cdot \ell'$ first splits the string $s$ into $s_1$ and $s_2$,
applies $\ell$ and $\ell'$ to $s_1$ and $s_2$ to get $t_1$ and $t_2$
respectively, then concatenates $t_1$ and $t_2$ and returns $t_1 \cdot t_2$ as
the final result in the forward direction. $\ell \cdot \ell'$ operates
similarly in the backward direction, but with $s, s_1$ and $s_2$ substituted
for $t, t_1$ and $t_2$. The $\mathit{swap} \; (\ell, \ell')$ lens operates
like $\ell \cdot \ell'$, except that it swaps $t_1$ and $t_2$ before
concatenating the two for a final result of $t_2 \cdot t_1$ in the forward
direction. In the backward direction, $\mathit{swap}(\ell, \ell')$ first undoes
the swap, then proceeds as expected. The $\ell \; | \; \ell'$ lens
chooses to apply $\ell$ or $\ell'$ depending on whether the source
(resp. target) data is matched by $\ell$ or $\ell'$ in the forward (resp.
backward direction). The $\ell^*$ lens splits the string $s$ into strings $s_1,
\ldots, s_n$, applies $\ell$ to each $s_i$ to get $t_i$, and then concatenates
each of the $t_i$'s for a final result of $t_1 \cdot \ldots \cdot t_n$ in the
forward direction. $\ell^*$ operates similarly in the backward direction, but
with $s, s_i$ substituted for $t, t_i$.

@@

\begin{prooftree}
\AxiomC{$\ell : R \Leftrightarrow S$}
\AxiomC{$R \equiv^s R'$}
\AxiomC{$S \equiv^s S'$}
\TrinaryInfC{$\ell : R' \Leftrightarrow S'$}
\end{prooftree}

@@

\begin{figure}[t]
\begin{center}
\AxiomC{$s_1 \in \Sigma^*$}
\AxiomC{$s_2 \in \Sigma^*$}
\BinaryInfC{$\mathit{const} \; (s_1, s_t): s_1 \Leftrightarrow s_2$}
\DisplayProof
\hskip 1.5em
\AxiomC{$R$ is strongly unambiguous}
\UnaryInfC{$\mathit{id} \; (R): R \Leftrightarrow R$}
\DisplayProof
\end{center}

\begin{center}
\AxiomC{$\substack{\ell_1 : R_1 \Leftrightarrow S_1 \\ \ell_2 : R_2
\Leftrightarrow S_2}$}
\AxiomC{$\substack{R_1 \cdot^! R_2 \\S_1 \cdot^! S_2 }$}
\BinaryInfC{$\ell_1 \cdot \ell_2: R_1 \cdot R_2 \Leftrightarrow S_1 \cdot
S_2$}
\DisplayProof
\hskip 1.5em
\AxiomC{$\substack{\ell_1 : R_1 \Leftrightarrow S_1 \\ \ell_2 : R_2
\Leftrightarrow S_2}$}
\AxiomC{$\substack{R_1 \cdot^! R_2 \\ S_2 \cdot^! S_1}$}
\BinaryInfC{$\mathit{swap} \; (\ell_1, \ell_2): R_1 \cdot R_2
\Leftrightarrow S_2 \cdot S_1$}
\DisplayProof
\end{center}

\begin{center}
\AxiomC{$\ell_1 : R_1 \Leftrightarrow R_2$}
\AxiomC{$\ell_2 : R_2 \Leftrightarrow R_3$}
\BinaryInfC{$\ell_2 \circ \ell_1: R_1 \Leftrightarrow R_3$}
\DisplayProof
\hskip 1.5em
\AxiomC{$\substack{\ell_1 : R_1 \Leftrightarrow S_1 \\ \ell_2 : R_2
\Leftrightarrow S_2}$}
\AxiomC{$\substack{\mathcal{L}(R_1) \cap \mathcal{L}(R_2) =
\varnothing \\ \mathcal{L}(S_1) \cap \mathcal{L}(S_2) = \varnothing}$}
\BinaryInfC{$\ell_1 \sep \ell_2: R_1 \sep R_2 \Leftrightarrow S_1 \sep S_2$}
\DisplayProof
\end{center}

@@

\subsection{Syntax of QRE Lenses}
Having given a brief overview of the class of bijective lenses, we now introduce
the class of QRE lenses. The language of QRE lenses is given by following
grammar:
$$ q := \kw{lift}(\ell) \sep \kw{lquot}(q, \ell) \sep
\kw{rquot}(\ell, q) \sep q \cdot q' \sep (q \sep q') \sep q^* \sep q \circ q',$$
where $\ell$ ranges over bijective lenses.

@@

\iffalse
The difference between QRE lenses and Boomerang quotient lenses lies in the
typing judgements. More concretely, each QRE lens $q$ has a type $q : c
\Leftrightarrow c_1$ where $c, c_1$ are QREs. In contrast, the approach used to
type quotient lenses in Boomerang is to classify equivalences according to
whether they are or are not the equality relation. This type system is based on
two observations: first, that most quotient lenses originate as lifted basic
lenses, and therefore have types whose equivalence relations are both equality;
and second, that equality is preserved by many of the quotient lens
combinators. Foster et al also discuss a second possible approach to typing
quotient lenses, where equivalence relations are represented by rational
functions that induce them. While this second approach is more refined than the
first, Boomerang favours the first approach since the second appears to be too
expensive to be useful in practice~\cite{quotientlenses}.
\fi

>>>>

% FOR Regular Expression names
\newcommand{\re}[1]{\ensuremath{\mathtt{#1}}}
\newcommand{\kw}[1]{\ensuremath{\mathit{#1}}}
\newcommand{\codefont}[1]{\ensuremath{\mathsf{#1}}}
\newcommand{\project}[2]{\ensuremath{\kw{project} \; #1 \mapsto #2}}
\newcommand{\squash}[3]{\ensuremath{\kw{squash} \; #1 \rightarrow #2\; \kw{using} \; #3}}
\newcommand{\perm}[2]{\ensuremath{\kw{perm}\; (#1)\; \kw{with}\; #2}}
\newcommand{\normalize}[3]{\ensuremath{\kw{normalize} \; (#1, #2, #3)}}
\newcommand{\sep}{\ensuremath{\ | \ }}
\newcommand{\canonizer}{\ensuremath{\kw{canonizer}}}
\newcommand{\bibtex}{\textsc{Bib}\TeX{}}
\newcommand{\get}{\ensuremath{\kw{get}}}
\newcommand{\semicolon}{\ensuremath{\; ; \;}}
\newcommand{\lput}{\ensuremath{\kw{put}}}
\newcommand{\create}{\ensuremath{\kw{create}}}
\newcommand{\const}{\ensuremath{\kw{const}}}
\newcommand{\swap}{\ensuremath{\kw{swap}}}

\newcommand{\eqrel}[1]{\ensuremath{\equiv_{#1}}}

@@

\sam{OK I suppose it's a trick though I don't see an obvious solution. }

@@

(* allow any permutation of fields interspersed with arbitrary whitespace *)
let bib_fields = perm (label, bib_author, bib_title) with (project ("," . wsp+) $\to$ ",\n")
let bibtex = "@book{" . bib_fields . "}"
\end{lstlisting}
\caption{QRE specification of \bibtex{} records. }
\label{fig:example-qre}
\end{figure}
\subsection{QRE Lenses and QRE Lens Synthesis}
\label{sec:examplesynth}
At this point we already have a tool for synthesizing bijective lenses from a
pair of regular expressions and a set of example input-output pairs, as well as
the language of QREs for defining equivalence relations on regular expressions.
An obvious next step in simplifying the quotient lens programming is thus to use
the bijective lens synthesis procedure as a subroutine for a quotient lens
synthesis procedure. This new procedure should take two QREs and a set
of example input-output pairs as input, compute the canonical data formats on
both sides of the lens from the pair of QREs, map the example input-output pairs
to their canonical representatives to get a set of examples in canonical form,
and then invoke the bijective lens synthesis procedure on the canonical data
formats and the new set of examples that are in canonical form.

Indeed this basic idea is what motivates our definition of {\em QRE Lenses}.
Intuitively, just like Boomerang's quotient lenses, our QRE lenses are bijective
lenses with ``canonizers at the edges''---Figure~\ref{fig:attheedges} presents
the architecture of each lens.
\begin{figure}[t]
\centering
\includegraphics[width=0.7\textwidth]{canonizers-outside}
\caption{Quotient Lens with ``Canonizers at the Edges''
\bcp{Maybe we should go a step further and label things like $K(c)$,
 $W(c)$, Canonizer(c), etc.  And probably we should display K as a subset of
 W.  I can make these changes later.)}}
\label{fig:attheedges}
\end{figure}
Every QRE lens $q_1$ has a type $q_1: c_1 \Leftrightarrow c_2$ where $c_1$ and $c_2$
are QREs. In the get direction, a QRE lens $q_1: c_1 \Leftrightarrow c_2$ uses the
source QRE $c_1$ to compute a unique representative for the data modulo the
equivalence relation defined by $c_1$ and then applies the $\get$ function of a
bijective lens $\ell_1$ to this representative. In the backward direction, $q_1$
operates similarly, but using the view QRE $c_2$ and the $\lput$ function of
$\ell_1$.

Because the QREs $c_1$ and $c_2$ determine the internal formats for data after
canonization, and because the algorithm for synthesizing bijective lenses is
directed by these formats, $c_1$ and $c_2$ are all that is required to
synthesize QRE lenses end-to-end. Indeed, we structured QRE lenses as bijective
lenses with canonizers at the edges, precisely to facilitate the use of this
synthesis algorithm.

@@

@@

\bcp{A question about variable names: If we use $R$ and $S$ for plain
  regular expressions, it seems inconsistent to use $c$ for QREs.  (And even
  more inconsistent to use $c$ and $c'$ or $c$ and $c_1$!) I suggest $P$ and
  $Q$ for QREs (with $Q$ as the default when we are only talking about
  one).}\sam{I made the notation consistent by just using $c_1$ and $c_2$. I
  personally feel that it is better not to use P and Q for QREs since we use p
  and q for QRE lenses, but if you disagree I will make the change.}\bcp{I
  agree that this is a problem.  But maybe the right fix is to choose
  different metavariables for QRE lenses.  Or to use capital letters other
  than $P$ and $Q$ for QREs.  I don't like $c_1$ and $c_2$, both for the
  lower-case letter and for the subscripts---it's not at all consistent with
our conventions elsewhere.}

Given regular expressions $R, S$ and equivalence relations
$\sim_R$ and $\sim_S$ defined on $\mathcal{L}(R)$ and $\mathcal{L}(S)$
respectively, we\bcp{can we clarify that this is exactly Foster et al's
definition?}\sam{this is not Foster et al's definition. It's just a
loosening of the definition for bijective lens used in the Optician
paper}\bcp{OK, then we should say this}define
a {\em bijective quotient lens} $q :
R /{\sim_R} \Leftrightarrow S/{\sim_S}$ to be a pair of functions $q.\get :
\mathcal{L}(R) \longrightarrow \mathcal{L}(S)$ and $q.\lput : \mathcal{L}(S)
\longrightarrow \mathcal{L}(R)$ such that for all $r \in \mathcal{L}(R)$ and $s
\in \mathcal{L}(S)$, then $q.\lput(q_1.\get(r)) \sim_R r$ and
$q.\get(q_1.\lput(s)) \sim_R s$. Moreover, if $r \sim_R r'$, then $q.\get(r) =
q.\get(r')$, and if $s \sim_S s'$, then $q.\lput(s) = q.\lput(s')$.

In words, a bijective quotient lens $q$ from $R$ to $S$ modulo $\sim_R$
and $\sim_S$ is a pair of functions $q.\get$ and $q.\lput$ such that
$q.\get$ respects $\sim_R$ and $q.\lput$ respects $\sim_S$, and such that
the {\em get} and {\em put} functions lifted to $\mathcal{L}(R)/{\sim_R}$ and
$\mathcal{L}(S)/{\sim_S}$ are mutual inverses.

These laws are similar to the laws for bijective lenses, except that the
equality restriction in the bijective lens laws are loosened to allow for
equivalence relations. Also, the condition that $r \sim_R r'$ implies
$q.\get(r) = q.\get(r')$ ensures that the {\em get} function induced on
the equivalence classes if $\equiv_c$ is well-defined, and similarly for
the {\em put} function.

Now let $c_1$ and $c_2$ be QREs. Given a bijection $\ell : K(c_1)
\Leftrightarrow K(c_2)$, we may define a quotient lens $q : W(c_1)/\eqrel{c_1}
\Leftrightarrow W(c_2)/\eqrel{c_2}$ by $q.\get = \ell \circ \canonizer(c_1)
$ and $q.\lput = \ell_1^{-1} \circ \canonizer(c_2)$. In other words, the get
function of the $q$ first canonizes the source data using $\canonizer(c_1)$
and then applies $f$ to the result, while the put function of the $q$ first
canonizes the target data using $\canonizer(c_2)$ and the applies $f^{-1}$ to
the result. This is our approach in defining QRE lenses.

One potential pitfall with QRE lenses is that the combinators that
we define on them may not preserve this normal form. For example, the
composition $q_1 \; ; \; q_2$ of QRE lenses $q_1$ and $q_2$ may not be
equivalent to any quotient lens that has the canonical normal form of a
bijective lens with canonizers a the edges. Consequently, in order to guarantee
that every QRE lens $q_1$ is equivalent to a QRE lens $q_2$ which has this
canonical form, we need to identify a set $S$ of bijections that will ensure
that all QRE lenses is functionally equivalent to a QRE lens in canonical form.
Our choice is to use the class of \textit{bijective lenses}, which we studied in
\cite{optician} and has this desired property. The following part of the paper
briefly discusses bijective lenses.

\subsection{Bijective Lenses}
We define the set of \textit{bijective lenses} to be the set of bijections
between regular languages created using Boomerang lens combinators.
The syntax for the language of bijective lenses is given by
$$\ell_1 := \mathit{id} \; (R) \sep \const(s_1, s_2) \sep  \swap(\ell_1,
\ell_2) \sep \ell_1 \cdot \ell_2 \; |  \; (\ell_1 \sep \ell_2) \sep \ell^* \;
| \; \ell_1 \; ; \;  \ell_2,$$ where $R$ ranges over regular expressions and $s$
ranges over character strings.

The denotation of a lens $\ell_1$ is $\llbracket \ell_1 \rrbracket \subseteq
\mathit{String} \times \mathit{String}$. If $(s_1, s_2) \in \llbracket \ell_1
\rrbracket$, then $\ell_1$ maps between $s_1$ and $s_2$.
\begin{align*}
\llbracket R \rrbracket &= \{(r, r) \sep r \in \mathcal{L}(R)\}\\
\llbracket \const(r, s) \rrbracket &= \{(r, s)\}\\
\llbracket \swap(\ell_1, \ell_2) \rrbracket &= \{(s \cdot t, t' \cdot s') \sep
(s, s') \in \llbracket \ell_1 \rrbracket \text{ and } (t, t') \in \llbracket
\ell_2 \rrbracket\}\\
\llbracket \ell_1 \cdot \ell_2 \rrbracket &= \{(s \cdot t, s' \cdot t) \sep
(s, s') \in \llbracket \ell_1 \rrbracket \text{ and } (t, t') \in \llbracket
\ell_2 \rrbracket\}\\
\llbracket \ell_1 \sep \ell_2 \rrbracket &= \{(s \cdot t) \sep
(s, t) \in \llbracket \ell_1 \rrbracket \text{ or } (s, t) \in \llbracket
\ell_2 \rrbracket\}\\
\llbracket \ell^* \rrbracket &= \{(s_1 \cdot \ldots \cdot s_n, t_1 \cdot \ldots
\cdot t_n) \sep (s_i, t_i) \in \llbracket \ell \rrbracket \text{ for } 1
\leq i \leq n\}\\
\llbracket \ell_1 \; ; \; \ell_2 \rrbracket &= \{(r, t) \sep \text{there exists }s
\text{ such that } (r, s) \in \llbracket \ell_1 \rrbracket \text{ and } (s, t)
\in \llbracket \ell_2 \rrbracket\}
\end{align*}
The $\mathit{\const}(s, t)$ lens replaces the string $s$ with $t$ in the source
data in the forward direction, and $t$ with $s$ in the target data in the backward
direction. $\mathit{id}(R)$ applies the identity function to the source and target
in $\mathcal{L}(R)$ in both directions. The composition lens $\ell_1 \; ; \; \ell_2$
applies $\ell_1$ followed by then $\ell_2$ to the source in the forward direction,
and applies $\ell_2$ followed by $\ell_1$ to the target in the backward direction.
The lens $\ell_1 \cdot \ell_2$ first splits the string $s$ into $s_1$ and $s_2$,
applies $\ell_1$ and $\ell_2$ to $s_1$ and $s_2$ to get $t_1$ and $t_2$
respectively, then concatenates $t_1$ and $t_2$ and returns $t_1 \cdot t_2$ as
the final result in the forward direction. $\ell_1 \cdot \ell_2$ operates
similarly in the backward direction, but with $s, s_1$ and $s_2$ substituted
for $t, t_1$ and $t_2$. The $\mathit{\swap} \; (\ell_1, \ell_2)$ lens operates
like $\ell_1 \cdot \ell_2$, except that it swaps $t_1$ and $t_2$ before
concatenating the two for a final result of $t_2 \cdot t_1$ in the forward
direction. In the backward direction, $\mathit{\swap}(\ell_1, \ell_2)$ first undoes
the \swap, then proceeds as expected. The $\ell_1 \; | \; \ell_2$ lens
chooses to apply $\ell_1$ or $\ell_2$ depending on whether the source
(resp. target) data is matched by $\ell_1$ or $\ell_2$ in the forward (resp.
backward direction). The $\ell^*$ lens splits the string $s$ into strings $s_1,
\ldots, s_n$, applies $\ell_1$ to each $s_i$ to get $t_i$, and then concatenates
each of the $t_i$'s for a final result of $t_1 \cdot \ldots \cdot t_n$ in the
forward direction. $\ell^*$ operates similarly in the backward direction, but
with $s, s_i$ substituted for $t, t_i$.

@@

\begin{prooftree}
\AxiomC{$\ell_1 : R \Leftrightarrow S$}
\AxiomC{$R \equiv^s R'$}
\AxiomC{$S \equiv^s S'$}
\TrinaryInfC{$\ell_1 : R' \Leftrightarrow S'$}
\end{prooftree}

@@

\begin{figure}[t]
\begin{center}
\AxiomC{$s_1 \in \Sigma^*$}
\AxiomC{$s_2 \in \Sigma^*$}
\BinaryInfC{$\mathit{\const} \; (s_1, s_t): s_1 \Leftrightarrow s_2$}
\DisplayProof
\hskip 1.5em
\AxiomC{$R$ is strongly unambiguous}
\UnaryInfC{$\mathit{id} \; (R): R \Leftrightarrow R$}
\DisplayProof
\end{center}

\begin{center}
\AxiomC{$\substack{\ell_1 : R_1 \Leftrightarrow S_1 \\ \ell_2 : R_2
\Leftrightarrow S_2}$}
\AxiomC{$\substack{R_1 \cdot^! R_2 \\S_1 \cdot^! S_2 }$}
\BinaryInfC{$\ell_1 \cdot \ell_2: R_1 \cdot R_2 \Leftrightarrow S_1 \cdot
S_2$}
\DisplayProof
\hskip 1.5em
\AxiomC{$\substack{\ell_1 : R_1 \Leftrightarrow S_1 \\ \ell_2 : R_2
\Leftrightarrow S_2}$}
\AxiomC{$\substack{R_1 \cdot^! R_2 \\ S_2 \cdot^! S_1}$}
\BinaryInfC{$\mathit{\swap} \; (\ell_1, \ell_2): R_1 \cdot R_2
\Leftrightarrow S_2 \cdot S_1$}
\DisplayProof
\end{center}

\begin{center}
\AxiomC{$\ell_1 : R_1 \Leftrightarrow R_2$}
\AxiomC{$\ell_2 : R_2 \Leftrightarrow R_3$}
\BinaryInfC{$\ell_1 \; ; \; \ell_2: R_1 \Leftrightarrow R_3$}
\DisplayProof
\hskip 1.5em
\AxiomC{$\substack{\ell_1 : R_1 \Leftrightarrow S_1 \\ \ell_2 : R_2
\Leftrightarrow S_2}$}
\AxiomC{$\substack{\mathcal{L}(R_1) \cap \mathcal{L}(R_2) =
\varnothing \\ \mathcal{L}(S_1) \cap \mathcal{L}(S_2) = \varnothing}$}
\BinaryInfC{$\ell_1 \sep \ell_2: R_1 \sep R_2 \Leftrightarrow S_1 \sep S_2$}
\DisplayProof
\end{center}

@@

\subsection{Syntax of QRE Lenses}
Having given a brief overview of the class of bijective lenses, we now introduce
the class of QRE lenses. The language of QRE lenses is given by following
grammar:
$$ q := \kw{lift}(\ell) \sep \kw{lquot}(q, \ell) \sep
\kw{rquot}(\ell, q) \sep q_1 \cdot q_2 \sep (q_1 \sep q_2) \sep q^* \sep
q_1 \; ; \; q_2,$$ where $\ell$ ranges over bijective lenses.

@@

\iffalse
The difference between QRE lenses and Boomerang quotient lenses lies in the
typing judgements. More concretely, each QRE lens $q_1$ has a type $q_1 : c
\Leftrightarrow c_1$ where $c, c_1$ are QREs. In contrast, the approach used to
type quotient lenses in Boomerang is to classify equivalences according to
whether they are or are not the equality relation. This type system is based on
two observations: first, that most quotient lenses originate as lifted basic
lenses, and therefore have types whose equivalence relations are both equality;
and second, that equality is preserved by many of the quotient lens
combinators. Foster et al also discuss a second possible approach to typing
quotient lenses, where equivalence relations are represented by rational
functions that induce them. While this second approach is more refined than the
first, Boomerang favours the first approach since the second appears to be too
expensive to be useful in practice~\cite{quotientlenses}.
\fi
